### 2017-11-06 Draft2/Appending.Dynamic.tex
Reach the bookmark

[make corollary]
If $\VarCapacitySeq_i$ exists and $i \geq 1$, then clearly $\FuncGrow$ must have been called when the size was $\VarCapacitySeq_{i - 1}$. Then the \textbf{growth sequence}, the sizes for which $\FuncGrow$ is called when $n$ items are appended, is $\VarCapacitySeq$ with the last term removed:
\begin{align*}
\VarGrowSeq = \VarCapacitySeq \setminus \left\{ \VarCapacitySeq_\VarUseful \right\}
\end{align*}

When $\FuncGrow$ is called and the current size is $\VarGrowSeq_i$, the algorithm copies $\VarCapacitySeq_i$ items. Then the total number of items copied when $n$ items are appended is:
\begin{align*}
\sum_i {\VarGrowSeq_i} &= \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \ldots + \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity\\
&= \left( \frac{\VarGrowthFactor^{\VarUseful} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Counting the writes made for each item by $\FuncAppend$, an explicit formula for $\FuncWrites(n)$ is as follows:
[ref1]
\begin{align*}
\FuncWrites(n) = n + \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Now, my goal is to approximate $\FuncWrites(n)$ with $\sim$. To make is easier to do so, I will asymptotically bound $\VarGrowthFactor^\VarUseful$ which depends on $n$.

>>>>

\begin{corollary}
\label{coro:GrowthSeq}
	Let the \textbf{growth sequence}, $\VarGrowSeq$, of $\VarList$ be the sizes for which $\FuncGrow$ is called when $n$ items are appended. Then $\VarGrowSeq = \VarCapacitySeq \setminus \left\{ \VarCapacitySeq_\VarUseful \right\}$.
\end{corollary}

\begin{proof}
	If $\VarCapacitySeq_i$ exists and $i \geq 1$, then clearly $\FuncGrow$ must have been called when the size was $\VarCapacitySeq_{i - 1}$, so $\VarCapacitySeq_{i - 1} \in \VarGrowSeq$. Then $\VarGrowSeq$ contains every term in $\VarCapacitySeq$ except for the last, $\VarCapacitySeq_\VarUseful$, as stated by the corollary.
\end{proof}
[make corollary]
If $\VarCapacitySeq_i$ exists and $i \geq 1$, then clearly $\FuncGrow$ must have been called when the size was $\VarCapacitySeq_{i - 1}$. Then the \textbf{growth sequence}, the sizes for which $\FuncGrow$ is called when $n$ items are appended, is $\VarCapacitySeq$ with the last term removed:
\begin{align*}
\VarGrowSeq = \VarCapacitySeq \setminus \left\{ \VarCapacitySeq_\VarUseful \right\}
\end{align*}

When $\FuncGrow$ is called and the current size is $\VarGrowSeq_i$, the algorithm copies $\VarGrowSeq_i$ items. Then the total number of items copied when $n$ items are appended is:
\begin{align*}
\sum_i {\VarGrowSeq_i} &= \VarInitCapacity + \VarGrowthFactor\VarInitCapacity + \ldots + \VarGrowthFactor^{\VarUseful - 1}\VarInitCapacity\\
&= \left( \frac{\VarGrowthFactor^{\VarUseful} - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Counting the writes made for each item by $\FuncAppend$, an explicit formula for $\FuncWrites(n)$ is as follows:
[ref1]
\begin{align*}
\FuncWrites(n) = n + \left( \frac{\VarGrowthFactor^\VarUseful - 1}{\VarGrowthFactor - 1} \right) \VarInitCapacity
\end{align*}
Now, my goal is to approximate $\FuncWrites(n)$ with $\sim$. To make is easier to do so, I will asymptotically bound $\VarGrowthFactor^\VarUseful$ which depends on $n$.

### 2017-11-06 Draft2/Appending.Growth.tex
Reach the bookmark

I again start off by finding the write cost for $n$ items. \ref{lem:CapacitySeq} still holds, since all of the assumptions made there still hold for growth arrays. In particular, although growth arrays use a different growth algorithm than dynamic arrays, the following claim is still true:

\begin{lemma}
\label{lem:GrowthArraysGrowthFactor}
	The capacity of a non-empty growth array grows by the constant factor $\VarGrowthFactor$.
\end{lemma}

\begin{proof}
	I prove that the $\FuncGrow$ algorithm always enforces this using induction. I induct on the number of times $\FuncGrow$ is called, $k$, showing that for all $k \geq 2$, $\FuncGrow$ behaves correctly when called the $k$th time. (At $k = 1$, the growth array is empty, so its behavior is not relevant to this lemma.) I will let $c_i(k)$ and $c_f(k)$ denote the initial/final capacities and $h_i(k)$ and $h_f(k)$ denote the initial/final head capacities for the $k$th call, respectively.

	For $k = 2$, $c_i(2) = \VarInitCapacity$. I wish to show that $c_f(2) = \VarGrowthFactor\VarInitCapacity$. This happens if and only if the next buffer has size ${\Delta c}(2) = (\VarGrowthFactor - 1)\VarInitCapacity$, which the algorithm ensures.

	For $k > 2$, by induction $c_i(k) = c_f(k - 1) = \VarGrowthFactor^{k - 2}\VarInitCapacity$ and $h_i(k) = h_f(k - 1) = (\VarGrowthFactor^{k - 2} - \VarGrowthFactor^{k - 3})\VarInitCapacity$. I wish to show $c_f(k) = \VarGrowthFactor^{k - 1}\VarInitCapacity$ and $h_f(k) = (\VarGrowthFactor^{k - 1} - \VarGrowthFactor^{k - 2})\VarInitCapacity$. Because $k > 2$, the algorithm will calculate $h_f(k)$ as $\VarGrowthFactor$ times $h_i(k)$. Then
	\begin{align*}
	h_f(k) &= \VarGrowthFactor h_i(k)\\
	&= \VarGrowthFactor (\VarGrowthFactor^{k - 2} - \VarGrowthFactor^{k - 3})\VarInitCapacity\\
	&= (\VarGrowthFactor^{k - 1} - \VarGrowthFactor^{k - 2})\VarInitCapacity
	\end{align*}
	and
	\begin{align*}
	c_f(k) &= c_i(k) + h_f(k)\\
	&= \VarGrowthFactor^{k - 2}\VarInitCapacity + (\VarGrowthFactor^{k - 1} - \VarGrowthFactor^{k - 2})\VarInitCapacity\\
	&= \VarGrowthFactor^{k - 1}\VarInitCapacity
	\end{align*}
	as desired.
\end{proof}

Now that \ref{lem:GrowthArraysGrowthFactor} and consequently \ref{lem:CapacitySeq} are proved for growth arrays, I examine the write cost of $\FuncGrow$. Unlike dynamic arrays, $\FuncGrow$ does not make $s$ writes if the current size is $s$. In fact, no items are copied; the only source of writes is when a buffer is appended to the tail, since the tail is a dynamic array.

Let $\FuncWritesByGrow(n)$ denote the total number of writes made by $\FuncGrow$, and let $\VarNumItemsTail$ be the size of the tail. Consider that [make corollary] also holds true for growth arrays, so $\FuncGrow$ is called $|\VarGrowSeq|$ times. A buffer is appended to the tail every time except the first, so the size of the tail is
\begin{align*}
\VarNumItemsTail &= |\VarGrowSeq| - 1 = |\VarCapacitySeq| - 2 = \VarUseful
\end{align*}
Then the formula for $\FuncWritesByGrow(n)$ is simply $\hat{\FuncWrites}(\VarUseful)$, where $\hat{\FuncWrites}$ is the write cost function for dynamic arrays. Finally, adding the writes made by $\FuncAppend$, the formula for $\FuncWrites(n)$ is
\begin{align*}
\FuncWrites(n) = n + \hat{\FuncWrites}(\VarUseful)
\end{align*}

>>>>

I start off again by finding the write cost for $n$ items. Lemma \ref{lem:CapacitySeq} still holds, since growth arrays satisfy the properties used by that proof. In particular, although growth arrays use a different growth algorithm than dynamic arrays, the following claim is still true:

\begin{lemma}
\label{lem:GrowthArraysGrowthFactor}
	The capacity of a growth array grows by the constant factor $\VarGrowthFactor$.
\end{lemma}

\begin{proof}
	I prove that the $\FuncGrow$ algorithm enforces this using induction. I induct on the number of times $\FuncGrow$ is called, $k$, showing that for all natural numbers $k$, $\FuncGrow$ behaves correctly when called the $k$th time. I will let $c_i$ and $c_f$ denote the initial/final capacities and $h_i$ and $h_f$ denote the initial/final head capacities for the $k$th call, respectively.

	For $k = 1$, $c_i = \VarInitCapacity$. I wish to show that $c_f = \VarGrowthFactor\VarInitCapacity$. This happens if and only if the next buffer has size $\Delta c = (\VarGrowthFactor - 1)\VarInitCapacity$, which the algorithm ensures.

	For $k > 1$, by induction $c_i = \text{previous }c_f = \VarGrowthFactor^{k - 2}\VarInitCapacity$, and $h_i = \text{previous }h_f = (\VarGrowthFactor^{k - 2} - \VarGrowthFactor^{k - 3})\VarInitCapacity$. I wish to show $c_f = \VarGrowthFactor^{k - 1}\VarInitCapacity$ and $h_f = (\VarGrowthFactor^{k - 1} - \VarGrowthFactor^{k - 2})\VarInitCapacity$. Because $k > 1$, the algorithm will calculate $h_f$ as $\VarGrowthFactor$ times $h_i$. Then
	\begin{align*}
	h_f &= \VarGrowthFactor h_i = \VarGrowthFactor (\VarGrowthFactor^{k - 2} - \VarGrowthFactor^{k - 3})\VarInitCapacity = (\VarGrowthFactor^{k - 1} - \VarGrowthFactor^{k - 2})\VarInitCapacity
	\end{align*}
	and
	\begin{align*}
	c_f &= c_i + h_f = \VarGrowthFactor^{k - 2}\VarInitCapacity + (\VarGrowthFactor^{k - 1} - \VarGrowthFactor^{k - 2})\VarInitCapacity = \VarGrowthFactor^{k - 1}\VarInitCapacity
	\end{align*}
	as desired.
\end{proof}

Since Lemma \ref{lem:GrowthArraysGrowthFactor} has been proven, Lemma \ref{lem:CapacitySeq} and all results based on it must also hold true for growth arrays. Now, I am ready to find the write cost of $\FuncGrow$. Unlike dynamic arrays, $\FuncGrow$ does not make $\VarGrowSeq_i$ writes when the current size is $\VarGrowSeq_i$. In fact, $\FuncGrow$ does not copy \textit{any} items supplied by the user. Writes are only made when a buffer is appended to the tail, since the tail is a dynamic array.

Let $\FuncWritesByGrow(n)$ denote the total number of writes made by $\FuncGrow$, and let $\VarNumItemsTail$ be the size of the tail. Since Corollary \ref{coro:GrowthSeq} also holds true for growth arrays, $\FuncGrow$ is called $|\VarGrowSeq|$ times. A buffer is appended to the tail each time $\FuncGrow$ is called. Thus, the tail's size is
\begin{align*}
\VarNumItemsTail &= |\VarGrowSeq| = |\VarCapacitySeq| - 1 = \VarUseful
\end{align*}
Then the formula for $\FuncWritesByGrow(n)$ is simply $\FuncWritesTail(\VarUseful)$, where $\FuncWritesTail$ denotes the tail's write cost function, that is, the write cost function for dynamic arrays. Finally, adding the writes made by $\FuncAppend$, the formula for $\FuncWrites(n)$ is
\begin{align*}
\FuncWrites(n) = n + \FuncWritesTail(\VarUseful)
\end{align*}
Now, I approximate $\FuncWrites(n)$ using $\sim$. To do this, I will derive the big-O complexity of $\VarUseful$.

\begin{lemma}
	$\VarUseful = O(\log n)$.
\end{lemma}

\begin{proof}
	From \ref{lem:CapacitySeq}, $\VarUseful = \max(\ExprMessy, 0)$. As mentioned in Lemma \ref{lem:ToVarUsefulPowerBounds}, for sufficiently large $n$, $\VarUseful = \ExprMessy$. Then
	\begin{align*}
	\VarUseful \sim \ExprMessy \sim \left( \ExprMessyInner \right) \sim \log_{\VarGrowthFactor} n
	\end{align*}
	By [lemma], $O(\VarUseful) = O(\log_{\VarGrowthFactor} n) = O(\log n)$ as desired.
\end{proof}
